---
title: "708 HW7"
author: "Huiyue Li"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)
```


```{r}
library(tidyverse)
library(gsDesign)
```


### Question 1  
**(a)**  
(1)    
When $\exp(-\lambda m)=1/2$,   
then $-\lambda m=-\log(2)$, so $m=\log(2)/\lambda$  
(2)  
The hazard rates: $\lambda_A=\log(2)/17=0.041$, $\lambda_B=\log(2)/12=0.058$;   
The hazard ratio $\frac{\lambda_A}{\lambda_B}=\frac{m_B}{m_A}=\frac{12}{17}=0.706$.

<br>

**(b)**
$$H_0:\log\left\{\frac{\lambda_A}{\lambda_B} \right\}\ge0$$   
$$H_a:\log\left\{\frac{\lambda_A}{\lambda_B} \right\}<0$$

<br>

**(c)**
$d=\frac{(z_{\alpha}+z_\beta)^2}{\pi(1-\pi)\theta^2}=\frac{(1.96+1.04)^2}{0.5^2\times \left(\log(0.706)\right)^2}=297.03\approx298$
The number of deaths required is `298`.

<br>

**(d)**  
Using $d_1$ denotes the number of deaths in arm A, $d_2$ denotes the number of deaths in arm B.  
$d=d_1+d_2$ where $d_j=\frac{a}{2}\left[A-\frac{\exp(-\lambda_jL)}{\lambda_j} \left\{\exp(\lambda_jA)-1\right\}\right]$ where $j=1,2$. $a=10$ pts/month, $n=10A$.  

(i) $A=L$
```{r}
fun=function(A){
  lambda1=log(2)/17
  lambda2=log(2)/12
   d=298
   L=A
  d1=10/2*(A-exp(-lambda1*L)/lambda1*(exp(lambda1*A)-1))
  d2=10/2*(A-exp(-lambda2*L)/lambda2*(exp(lambda2*A)-1))
  d1+d2-d
}
10*uniroot(fun,c(0,100))$root
```
The number of patients to achieve the required number of deaths is `485`.  

<br>

(ii) The number of patients to achieve the required number of deaths is $n=1.5\times d=447$.  

<br>

(iii) $F=24$, $L=A+24$
```{r}
fun1=function(A){
  lambda1=log(2)/17
  lambda2=log(2)/12
  d=298
  L=A+24
  d1=10/2*(A-exp(-lambda1*L)/lambda1*(exp(lambda1*A)-1))
  d2=10/2*(A-exp(-lambda2*L)/lambda2*(exp(lambda2*A)-1))
  d1+d2-d
}
10*uniroot(fun1,c(0,100))$root
```
The number of patients to achieve the required number of deaths is `352`.   

I like the second design better, because compared with the first design, it requires a smaller sample size. And comparing with the third design, which has a long additional follow-up time and will spend much more time and money. But in real world, we usually cannot identify the exact ratio of $n$ and $d$ during designing the study.

<br>

**(e)**
The power of the study will decrease.   
We can adopt a conservative adjustment for sample size, which is $n_{adjusted}=n(1+0.1)$, where $n$ is the number of patients in **(d)**, `0.1` is the rate of withdrwal.  

<br>

### Question 2
**(a)**
```{r}
Po=gsDesign(k = 5,n.fix = 378,test.type = 2,alpha = 0.025,beta = 0.1,sfu = "Pocock")
Of=gsDesign(k = 5,n.fix = 378,test.type = 2,alpha = 0.025,beta = 0.1,sfu = "OF")
```
(1) For Pocock boundary, the critical values are `r {round(Po$upper$bound,3)}` and the maximum sample size is `r {ceiling(max(Po$n.I))}`.  
  
(2) For O’Brien-Fleming boundary, the critical values are `r {round(Of$upper$bound,3)}` and the maximum sample size is `r {ceiling(max(Of$n.I))}`.

<br>

**(b)**
(1) Pocock
```{r}
n.fix=378
rj=rep(0,5000)
set.seed(2022)
for (i in 1:5000) {
  X1=NULL
  X2=NULL
  t=NULL
  for (k in 1:5) {
    n=ifelse(k==1,ceiling(Po$n.I/2)[1],ceiling(Po$n.I/2)[k]-ceiling(Po$n.I/2)[k-1])
    X1=c(X1,rnorm(n,0,1))
    X2=c(X2,rnorm(n,0,1))
    t[k]=abs(mean(X1)-mean(X2))/sqrt(1/length(X1)+1/length(X2))
    if(t[k]>Po$upper$bound[k]) {rj[i]=1;break}
  }
}
```
The empirical type I error for Pocock boundary is `r {mean(rj)}`.

<br>

(2) O’Brien-Fleming
```{r}
n.fix=378
rj1=rep(0,5000)
set.seed(2022)
for (i in 1:5000) {
  X1=NULL
  X2=NULL
  t=NULL
  for (k in 1:5) {
    n=ifelse(k==1,ceiling(Of$n.I/2)[1],ceiling(Of$n.I/2)[k]-ceiling(Of$n.I/2)[k-1])
    X1=c(X1,rnorm(n,0,1))
    X2=c(X2,rnorm(n,0,1))
    t[k]=abs(mean(X1)-mean(X2))/sqrt(1/length(X1)+1/length(X2))
    if(t[k]>Of$upper$bound[k]) {rj1[i]=1;break}
  }
}
```
The empirical type I error for O’Brien-Fleming boundary is `r {mean(rj1)}`.
  
Given the above results, the type I errors are both controlled well by Pocock and O’Brien-Fleming boundaries since they are both smaller than 0.05.

<br>

**(c)**
(1) Pocock
```{r}
n.fix=378
Total=NULL
set.seed(2022)
for (i in 1:5000) {
  X1=NULL
  X2=NULL
  t=NULL
  for (k in 1:5) {
    n=ifelse(k==1,ceiling(Po$n.I/2)[1],ceiling(Po$n.I/2)[k]-ceiling(Po$n.I/2)[k-1])
    X1=c(X1,rnorm(n,1/3,1))
    X2=c(X2,rnorm(n,0,1))
    t[k]=abs(mean(X1)-mean(X2))/sqrt(1/length(X1)+1/length(X2))
    if(t[k]>Po$upper$bound[k]) {break}
  }
  Total[i]=length(X1)+length(X2)
}
```
The expected sample size for Pocock boundary is `r {ceiling(mean(Total))} `.

<br>

(2) O’Brien-Fleming
```{r}
n.fix=378
Total1=NULL
set.seed(2022)
for (i in 1:5000) {
  X1=NULL
  X2=NULL
  t=NULL
  for (k in 1:5) {
    n=ifelse(k==1,ceiling(Of$n.I/2)[1],ceiling(Of$n.I/2)[k]-ceiling(Of$n.I/2)[k-1])
    X1=c(X1,rnorm(n,1/3,1))
    X2=c(X2,rnorm(n,0,1))
    t[k]=abs(mean(X1)-mean(X2))/sqrt(1/length(X1)+1/length(X2))
    if(t[k]>Of$upper$bound[k]) {break}
  }
  Total1[i]=length(X1)+length(X2)
}
```
The expected sample size for O’Brien-Fleming boundary is `r {ceiling(mean(Total1))} `.

Given the above results, when using Pocock and O’Brien-Fleming designs, they could both reduce the expected sample size significantly under $H_a$, if comparing with the fixed sample design with sample size=378. Also, we can easily find that the expected sample size of Pocock design is smaller than which of the O’Brien-Fleming design under $H_a$.