---
title: "708 HW3"
author: "Huiyue Li"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = F,message = F)
```

## Question 1 Surrogate endpoint  
(1)	For each of the following situations, briefly explain why S may or may not be a good surrogate endpoint trial for Y.  
  
**A**. Surrogate  Outcome S is not a good surrogate endpoint for Clinical Outcome Y, since it is not in the casual pathway between intervention and true clinical outcome, i.e., there is no effect of intervention on S and S cannot predict Y.  
  
**B**. Surrogate  Outcome S is a good surrogate endpoint for Clinical Outcome Y, since it satisfied the related requirement that S lies the causal pathway between intervention Z and the true clinical outcome Y, i.e., the effect of the intervention Z on S predicts the effect on the true clinical outcome Y.  
  
**C**. Surrogate  Outcome S is not a good surrogate endpoint for Clinical Outcome Y, since it is not in the casual pathway between intervention and true clinical outcome, i.e., there is no effect of intervention on S.  

<br>

(2)  
The PE is `0.563`, which is much smaller than 1. And in model A, we can find that treatment Z does not have significant effect on surrogate outcome S; in model B, we can find that treatment Z does not have significant effect on true outcome T. Therefore, S is not a good surrogate endpoint for T.
```{r}
library(tidyverse)
armd=read.csv("armd.csv")
armd$Trt=ifelse(armd$Treat==1,1,0)
#Model A
Mol_A=lm(Diff24~Trt,armd)
summary(Mol_A)
#Model B
Mol_B=lm(Diff52~Trt,armd)
summary(Mol_B)
#Model C
Mol_C=lm(Diff52~Diff24,armd)
summary(Mol_C)
#Model D
Mol_D=lm(Diff52~Diff24+Trt,armd)
#PE
PE=round((Mol_B$coefficients[2]-Mol_D$coefficients[3])/Mol_B$coefficients[2],3)
names(PE)="PE"
PE
```

<br>

## Single group 1-stage and 2-stage designs:  
(1)  
Under $H_0$:
$$PET(p)=Pr(x_1 \le3)=\sum^{3}_{x_1=0}\binom{16}{x_1}0.15^{x_1}0.85^{16-x_1}=0.790$$
Under $H_A$:
$$PET(p)=Pr(x_1 \le3)=\sum^{3}_{x_1=0}\binom{16}{x_1}0.4^{x_1}0.6^{16-x_1}=0.065$$
```{r}
PET0=pbinom(3,16,0.15)
PETA=pbinom(3,16,0.4)
```


<br>

(2) 
The probability of rejecting $H_0$ (in stage 2) given $x_1$ (in stage 1) is $cp(x_1)=Pr(x_1+X_2>8|x_1)$. We sum $cp(x_1)$ over all possible $x_1 (>r_1; r_1=3)$ to obtain $Pr(reject \ H_0)$:  

Under $H_0$, the overall probability of rejection is
$$\rho(p_0)=\sum^{16}_{x_1=4}\left\{cp(x_1)\times binom(x_1,16,0.15)\right\}=0.034$$
Under $H_A$, the overall probability of rejection is
$$\rho(p_1)=\sum^{16}_{x_1=4}\left\{cp(x_1)\times binom(x_1,16,0.4)\right\}=0.903$$
The actual type I error is $\rho(p_0)=0.034$, and the actual type II error is $1-\rho(p_1)=0.097$.

```{r}
rho=function(p){
rho=0
for (x1 in 4:16){
rho1=(pbinom(16,16,p)-pbinom(8-x1,16,p))*dbinom(x1,16,p)
rho=rho+rho1
}
return(rho)
}

#p0
round(rho(0.15),3)
#p1
round(rho(0.4),3)
```




<br>

(3)  
The expected sample size for $p=0.15$  
$$EN(p=0.15)=n_1+n_2(1-PET(p=0.15))=16+16*(1-0.790)=19.36$$
The expected sample size for $p=0.4$  
$$EN(p=0.4)=n_1+n_2(1-PET(p=0.4))=16+16*(1-0.065)=30.96$$
```{r}
EN0=16*(2-0.79)
ENA=16*(2-0.065)
```

<br>

(4)   
Simon's optimal design is the design which with the smallest expected sample size under $H_0$. And Simon's minimax design is the design with the smallest maximal sample size n ($n=n_1+n_2$) inder $H_0$.  

<br>

(5)  
```{r}
library(clinfun)
cl=ph2simon(pu=0.15,pa=0.4,ep1=0.1,ep2=0.1)
cl
head(cl$out,10)
```
The Simon's optimal design is the 2-stage design ($r_1=1,n_1=10,r=5,n=22$), and the Simon's minimax design is the 2-stage design ($r_1=2,n_1=15,r=5,n=21$). The 2-stage design studied in (1) and (2) is not the Simon's optimal or minimax design. Because its expected sample size is 19.36>15.47, and its maximal sample size is 32>21.

<br>

(6)   
Type I error rate:  
$$Pr(reject H_0|H_0)=\sum^{32}_{x=9}\binom{32}{x}0.15^{x}0.85^{32-x}=0.041$$
Type II error rate:  
$$Pr(accept H_0|H_A)=1-\sum^{32}_{x=9}\binom{32}{x}0.4^{x}0.6^{32-x}=0.057$$
```{r}
st11=pbinom(32,32,0.15)-pbinom(8,32,0.15)
st12=1-pbinom(32,32,0.4)+pbinom(8,32,0.4)
```

<br>


(7) Given **(2)** and **(6)**, we can find that the Type I error and Type II error are close. However, under $H_0$ is true, the 2-stage design trial will frequently stop at stage 1 given $PET(p_0)=0.790$, then there are fewer subjects enrolling in the trial than 1-stage design. Whereas, under $H_A$ is true, the 2-stage design trial always has the chance to start stage 2 trial $PET(p_a)=0.065$. Overall, the 2-stage design is more likely to reduce the expected sample size, especially under the $H_0$ is true.